---
title       : Molecular Mass Calculator 
subtitle    : Shiny app created for Coursera Developing Data Products course    
author      : John Beal
job         : Chemist
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## Molecular Mass Calculator

[Relative molecular mass] (http://goldbook.iupac.org/R05271.html) is an important characteristic of molecular chemical compounds. It is defined as the ratio of the mass of a molecule to the [unified atomic mass unit] (http://goldbook.iupac.org/U06554.html). Its units are defined as \(\frac{1}{12}\)<sup>th</sup> the mass of a <sup>12</sup>C atom.

Calculating the molecular mass of a molecule from its molecular formula is a routine, but mundane task suitable for automation.

### CHNOPS

C - carbon, H - hydrogen, N - nitrogen, O - oxygen, P - phosphorus, S - sulfur

Combinations of these six chemical elements make up most biological molecules (and a significant subset of organic molecules too). For the sake of simplicity, the initial version of the molecular mass calculator is limited to these elements.  

---

## Methdology

$$Mr = \Sigma m_i.n_i$$

- \(Mr\), molecular mass. 
- \(mi\), atomic mass of element \(i\).
- \(ni\), number of atoms of element \(i\) in molecule.

Molecular mass is calculated as the sum of the product of the atomic mass of the element and number of atoms of the element in the molecule; for each element in the molecule. 

---

## Code ##

The algorithm is enacted in server.R by the following R code:

```{r}
## Atomic masses (i.e. coefficients for predicting molecular mass) for CHNOPS
atomicmass <- c(12.01, 1.008, 14.01, 16.00, 30.97, 32.07)
names(atomicmass) <- c("C","H","N","O", "P", "S")

## Function takes input of number of carbon (C), hydrogen (H), nitrogen (N), oxygen (O), 
## phosphorus (P) and sulfur (S) atoms in a molecule,and returns the molecular mass. 
## Function also requires vector of atomic masses.
molecularmass <- function (nC = 0, nH = 0, nN = 0, nO = 0, nP = 0, nS = 0, atomicmasses) {
                        round(sum(nC*atomicmasses["C"],
                                  nH*atomicmasses["H"],
                                  nN*atomicmasses["N"],
                                  nO*atomicmasses["O"],
                                  nP*atomicmasses["P"],
                                  nS*atomicmasses["S"]), 2)
}
```

---

## Example

Example for the important biological chemical ATP (adenosine triphosphate, C<sub>10</sub>H<sub>16</sub>N<sub>5</sub>O<sub>13</sub>P<sub>3</sub>):
inputting C = 10, H = 16, N = 5, O = 13, P = 3 returns the molecular mass of `r molecularmass(nC = 10, nH = 16, nN = 5, nO = 13, nP = 3, atomicmasses = atomicmass)`:

``` {r}
molecularmass(nC = 10, nH = 16, nN = 5, nO = 13, nP = 3, atomicmasses = atomicmass)
```









